---
title: "Usage examples for rstanmulticore"
author: "Nathan VanHoudnos" 
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage examples for rstanmulticore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rstanmulticore` is a cross-platform (Windows, Linux, and Mac) R package to parallelize RStan MCMC chains across multiple cores. The syntax is very simple: replace calls to stan(...) with pstan(...).

# Examples

We begin with the default "Eight Schools" example from the [Quick Start Guide](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#example-1-eight-schools) using the default `stan` function:

```{r}
library(rstan)

## The data to analyze (Yes, it is very little!)
schools_dat <- list(
  J = 8, y = c(28,  8, -3,  7, -1,  1, 18, 12),
  sigma = c(15, 10, 16, 11,  9, 11, 10, 18))

## The Stan model for the data, stored as a string
schools_code <- 'data {
  int<lower=0> J; // number of schools 
  real y[J]; // estimated treatment effects
  real<lower=0> sigma[J]; // s.e. of effect estimates 
}
parameters {
  real mu; 
  real<lower=0> tau;
  real eta[J];
}
transformed parameters {
  real theta[J];
  for (j in 1:J)
    theta[j] <- mu + tau * eta[j];
}
model {
  eta ~ normal(0, 1);
  y ~ normal(theta, sigma);
}'
      
## Estimating the model 
fit.serial   <- stan( model_code = schools_code, data = schools_dat, 
                      iter = 1000, chains = 4, seed = 1)
```


Note that `stan` is pretty verbose.

I chose to make `pstan` less verbose. By default, `pstan` reports sparse progress information to the R console and the more detailed information is redirected to a file, `stan-debug-*`, that is created in the current working directory. (If you wish to see the detailed info in real time, use `tail -f` in your shell.)

Usage of `pstan` as follows:

```{r}
library(rstanmulticore)

fit.parallel <- pstan( model_code = schools_code, data = schools_dat, 
                       iter = 1000, chains = 4, seed = 1)
```

Note that the serial `stan` fit and the parallel `pstan` fit give identical samples:
```{r}
all.equal( fit.serial@sim$samples, fit.parallel@sim$samples )
```

You can also pass in a `fit` object to save on compilation time:
```{r}
fit.parallel.2 <- pstan( fit = fit.serial, data = schools_dat,
                         iter = 1000, chains = 4, seed = 1)
all.equal( fit.serial@sim$samples, fit.parallel.2@sim$samples )
```

Or you use the `file` parameter from `stan`: 
```{r}
## First, create a txt file in current working directory containing
## the Stan code
getwd()
write(schools_code, file="schools_code.txt")

## Use the file to run pstan
fit.parallel.3 <- pstan( file = 'schools_code.txt', data = schools_dat,
                         iter = 1000, chains = 4, seed = 1)
all.equal( fit.serial@sim$samples, fit.parallel.3@sim$samples )
```

You can also pass in initial values to the chain
```{r}
  ## Set initial values
  chain1 <- list(mu = -100, tau = 1, eta = rep(100, 8))
  chain2 <- list(mu = 100, tau = 1, eta = rep(200, 8))
  chain3 <- list(mu = 1000, tau = 100, eta = rep(300.5, 8))
  chain4 <- list(mu = -1000, tau = 100, eta = rep(400, 8))

fit.serial.init   <-  stan( fit = fit.serial, data = schools_dat,
                            init = list( chain1, chain2, chain3, chain4),
                            iter = 10000, chains = 4, seed = 1)

fit.parallel.init <- pstan( fit = fit.serial, data = schools_dat,
                            init = list( chain1, chain2, chain3, chain4),
                            iter = 10000, chains = 4, seed = 1)

all.equal( fit.serial.init@sim$samples, fit.parallel.init@sim$samples )
```
